Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-07-06 WSSSPE2_report/WSSSPE2_review.tex
bib style is vancouver now (JSOR's)
fix typo
lots of little things in response to previous changes and other fixes

\begin{abstract}
This technical report records and discusses the Second Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE2). 
%The workshop used an
%alternative submission and peer-review process, which led to a set of papers
%divided across five topic areas: 
The report includes a description of the alternative, experimental submission
and review process, two workshop keynote presentations, a series of lightning
talks, a discussion on sustainability, and five discussions from the topic areas
of exploring sustainability; software development experiences; credit \&
incentives; reproducibility \& reuse \& sharing; and code testing \& code
review. For each topic, the report includes a list of tangible actions that were
proposed and would lead to potential changes.

First and foremost, the workshop recognized that reliance on scientific software
is pervasive in all areas of world-leading research today. Then the workshop
participants proceeded to explore different perspectives on the concept of
sustainability. Key enablers and barriers of sustainable scientific software
were identified from the experiences of the workshop participants. In addition,
recommendations with new requirements such as software credit files and software
prize frameworks were outlined for improving practices in sustainable software
engineering.

There was also broad consensus that, within the disciplines represented, formal
training in software development or engineering was rare among the
practitioners. Significant strides are needed to be made in building a sense of
community via training in software and technical practices, on increasing their
size and scope, and on better integrating them directly into graduate education
programs.

Journals can define and publish policies to improve reproducibility, whereas
reviewers can insist that authors provide sufficient information and access to
data and software to allow them reproduce the results in the paper. Hence a list of
critieria is compiled for  journals to provide to reviewers so as to make it easier to
review software submitted for publication as a ``Software Paper.''

@@

\item \textbf{Marlon Pierce: Patching It Up, Pulling It
Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
software needs a diverse, openly governed community behind it, just as it needs
open licensing and a public code repository. To probe the level of governance
within open source projects, he and his collaborators %(S.~Marru and C.~Mattmann)
suggested a contest to encourage individual developers to submit
patches and requests to projects that are important to them. This simple
mechanism shall expose several governance mechanisms, such as how easy it is for
independent developers to communicate with project leadership, how projects
accept and license third-party contributions, and how projects make
decisions such as granting source tree write access.

@@

\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
Howison sought to provoke discussion by proposing that papers whose workflows
are not kept current with the changing software ecosystem should be
automatically retracted. This would create an incentive for authors to keep
their software current and usable, rather than the current situation in which
every potential user has to do this individually. A softer version of the
proposal would identify papers whose software workflow has become bit-rotten and
allow others to keep the code up to date, either adding them as new authors of
the paper or providing credit in some other form. %for their academic service.

@@

The general responses to what software sustainability meant were:
\begin{itemize}
\item keeping software scientifically useful
\item separating techniques in code from knowledge in code
\item that an adequately large community finds value in software and is willing
to sustain it.
\end{itemize}
\katznote{any discussion of this?}

The enablers of and barriers to software sustainability, roughly ranked by
attendee votes are shown in
{\tablename}s~\ref{tb:software_sustainability_enablers} and
\ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other
items were suggested as barriers, but were not voted on due to lack of time in
the session:
layering up dependencies;
using software past its sustainable life;
using software past its usable life;
inertia for accepted answers versus wrong or right answers;
monolithic or poor code; and
need to restructure code when hardware/software/libraries change.
}
\katznote{any discussion of these tables?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Software Development Experiences} \label{sec:devel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Nancy led writing of this section, with Subgroup A part from Matt Jones}

\note{\href{http://tinyurl.com/pn4eq8z}{Google doc notes}}
%https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit

@@

\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Credit \& Incentives} \label{sec:credit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Dan led writing of this section}

\note{\href{http://tinyurl.com/k8ruyn9}{Google doc notes}}
%https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q

@@

\note{for reference, notes are in \url{http://tinyurl.com/l4tc6a4}}
%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit

This group defined its goal as creating a low-barrier-of-entry method, to be called a credit file,  for
recording names and roles (and in a second phase, optionally including weights)
of contributors (coders and other intellectual contributions) to a software
package in a machine readable way, then encouraging
the scientific community to adopt this practice.

@@

\subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
\note{by Frank Seinstra}

\note{for reference, notes are in \url{http://tinyurl.com/mx58tmx}}
%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit

@@

\subsubsection{Remaining breakout discussions: Prizes}\label{sec:prizes}
\note{sub-breakout notes: \url{http://tinyurl.com/mjok4o5}}
%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit

@@

Other (non-prize) forms of incentives are
\begin{itemize}
\item Converting reputation by joining the ASF, Google, etc.
\item Inviting ASF and open source people to contribute to scientific code (at
least on overlapping questions)
\item Template for assessing scientific contributions made through software.
\end{itemize}

@@

One of the groups's ideas was to create a funding program for disciplines or
other organizations to create a prize program. We would provide a framework and
a set of requirements, for example: awards to individuals; must award in 5 or 6
areas; must have objective criteria and a jury that includes senior/junior
domain experts and technologists; must have the recipients awarded at a
relevant event; and must provide citations that explain why the prizes should be
awarded. Different organizations could then decide to sign up to the framework
and give out awards under this general brand. However, there was a concern that
having many organizations award prizes may reduce the impact of the prizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reproducibility \& Reuse \& Sharing} \label{sec:reproduce}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Neil led writing of this section}

\note{\href{http://tinyurl.com/kqpe87z}{Google doc notes}}
%https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9

@@

The first group discussed ways in which reproducibility of papers could be
improved. A consensus surrounded the provision of examples: demonstrating to
others how to achieve reproducibility. Major public funding investments go into
research that heavily relies on reproducible software, hence lack thereof
is raising major concerns.

@@

Other discussion focused on ways in which researchers themselves could improve
the reproducibility of research. One way would be to establish tracks at
conferences that subject papers to reproduction, which for those that pass would
lend them an additional badge (similar to OOPSLA). Another is simply to get more
people to use your software: for instance by outreach to high school students ---
can your software be used by them? Finally, there is a role for
community-curated benchmarks to validate the performance and capabilities of
tools.

@@

An issue around the sustainability of software catalogues is that their
usefulness often depends on the domain. For instance, in the bio-sciences, there
are more homogeneous data and standard shred code. In areas like ecology,
code is often specific to a problem, meaning that the level of re-use might
be at a general statistical level of abstraction, but then every research
use is highly customized.

@@

\item Paper Policies
\begin{itemize}
	\item Required sections
	\item Keywords
	\begin{itemize}
		\item Paper provides keywords to help describe software
	\end{itemize}
	\item Papers can be updated when new releases of software are available
	\begin{itemize}
		\item At no extra cost/at significantly reduced cost
	\end{itemize}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code Testing \& Code Review} \label{sec:code_testing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Dan leading writing of this section, with help from Tom Clune and Karen Cranston}

\note{\href{http://tinyurl.com/l5t5h45}{Google doc notes}}
%https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur

@@

\item Formal tutorials at conferences (for example, a test driven development
(TDD) at a Software Developers Best Practices conference): These often require
self-learning following a tutorial, but it can be hard to find a tutorial that
matches the programming language and the domain of the participant.

@@

Teach testing by beginning with a (smallish) piece of code that lacks
appropriate tests and develop an exercise that involves refactoring the code
into a more presentable form through creating/deriving unit tests. Students
would extract unit-tests through reverse-engineering or questioning a
partner who is a developer/expert.

This leads to a more general discussion about whether programming courses
should be part of the standard curriculum for science students and the
challenges of fitting additional material into university curricula. This is the
reasons that short workshops such as Software Carpentry and others exist.

@@

\subsubsection{Writing tests for scientific software}
The second major discussion thread in the group addressed the issues identified
by Software Carpentry --- the difficulties in writing tests appropriate for
scientific software: %Even among this group, while many 
What tests are
appropriate to ensure that an complex method is working correctly? When is the
result of a numerical computation `close enough' to pass a test?

Several members of the group indicated that while they valued the concept of
software testing in theory, they were unaware of how to test certain kinds of
software. Or, more importantly, they lacked relevant examples of software
testing that would be suitable for members new to their teams. As an example, the
group asked if it could produce a 100-line example of how to test neutron
transport (or other specific scientific examples) targeted at a sophomore? This
would be a demonstration of ``How do I do this right?''

@@

The fact remains that contributors want to get credit for their
participation in the process. The workshop organizers also want to make
sure that the workshop content and their efforts are recorded. Ideally, there
would be a service that could index the contributions to the
workshop, serving the authors, the organizers, and the larger community. 
Since there still isn't such a service today, the workshop organizers are
writing this initial report and making use of arXiv as a partial solution to
provide a record of the workshop.

@@

\todo{feel free to add stuff here}

@@

\bibliographystyle{plain}

>>>>

\begin{abstract}
This technical report records and discusses the Second Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE2). 
%The workshop used an
%alternative submission and peer-review process, which led to a set of papers
%divided across five topic areas: 
The report includes a description of the alternative, experimental submission
and review process, two workshop keynote presentations, a series of lightning
talks, a discussion on sustainability, and five discussions from the topic areas
of exploring sustainability; software development experiences; credit \&
incentives; reproducibility \& reuse \& sharing; and code testing \& code
review. For each topic, the report includes a list of tangible actions that were
proposed and that would lead to potential change.

First and foremost, the workshop recognized that reliance on scientific software
is pervasive in all areas of world-leading research today. The workshop
participants then proceeded to explore different perspectives on the concept of
sustainability. Key enablers and barriers of sustainable scientific software
were identified from their experiences. In addition,
recommendations with new requirements such as software credit files and software
prize frameworks were outlined for improving practices in sustainable software
engineering.

There was also broad consensus that, within the disciplines represented, formal
training in software development or engineering was rare among the
practitioners. Significant strides need to be made in building a sense of
community via training in software and technical practices, on increasing their
size and scope, and on better integrating them directly into graduate education
programs.

Journals can define and publish policies to improve reproducibility, whereas
reviewers can insist that authors provide sufficient information and access to
data and software to allow them reproduce the results in the paper. Hence a list of
criteria is compiled for  journals to provide to reviewers so as to make it easier to
review software submitted for publication as a ``Software Paper.''

@@

\item \textbf{Marlon Pierce: Patching It Up, Pulling It
Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
software needs a diverse, openly governed community behind it, just as it needs
open licensing and a public code repository. To probe the level of governance
within open source projects, he and his co-authors (Marru and Mattmann)
suggested a contest to encourage individual developers to submit
patches and requests to projects that are important to them. This simple
mechanism shall expose several governance mechanisms, such as how easy it is for
independent developers to communicate with project leadership, how projects
accept and license third-party contributions, and how projects make
decisions such as granting source tree write access.

@@

\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
Howison sought to provoke discussion by proposing that papers whose workflows
are not kept current with the changing software ecosystem should be
automatically retracted. This would create an incentive for authors to keep
their software current and usable, rather than the current situation in which
every potential user has to do this individually. A softer version of the
proposal would identify papers whose software workflow has become bit-rotten and
allow others to keep the code up to date, either adding them as new authors of
the paper or providing credit for their academic service in some other form.

@@

The general responses to what software sustainability meant were:
\begin{itemize}
\item keeping software scientifically useful
\item separating techniques in code from knowledge in code
\item that an adequately large community finds value in software and is willing
to sustain it.
\end{itemize}
%\katznote{any discussion of this?}

The enablers of and barriers to software sustainability, roughly ranked by
attendee votes are shown in
{\tablename}s~\ref{tb:software_sustainability_enablers} and
\ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other
items were suggested as barriers, but were not voted on due to lack of time in
the session:
layering up dependencies;
using software past its sustainable life;
using software past its usable life;
inertia for accepted answers versus wrong or right answers;
monolithic or poor code; and
need to restructure code when hardware/software/libraries change.
}
%\katznote{any discussion of these tables?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Software Development Experiences} \label{sec:devel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Nancy led writing of this section, with Subgroup A part from Matt Jones}

%\note{\href{http://tinyurl.com/pn4eq8z}{Google doc notes}}
%https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit

@@

%\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Credit \& Incentives} \label{sec:credit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Dan led writing of this section}

%\note{\href{http://tinyurl.com/k8ruyn9}{Google doc notes}}
%https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q

@@

%\note{for reference, notes are in \url{http://tinyurl.com/l4tc6a4}}
%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit

This group defined its goal as creating a low-barrier-of-entry method for
recording names and roles (and in a second phase, optionally including weights)
of contributors (coders and other intellectual contributions) to a software
package in a machine readable way (to be called a credit file), then encouraging
the scientific community to adopt this practice.

@@

\subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
%\note{by Frank Seinstra}

%\note{for reference, notes are in \url{http://tinyurl.com/mx58tmx}}
%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit

@@

\subsubsection{Remaining breakout discussions: Prizes}\label{sec:prizes}
%\note{sub-breakout notes: \url{http://tinyurl.com/mjok4o5}}
%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit

@@

Other (non-prize) forms of incentives are
\begin{itemize}
\item Converting reputation by joining the ASF, Google, etc.
\item Inviting ASF and open source people to contribute to scientific code (at
least in areas where there is overlapping interest).
\item Template for assessing scientific contributions made through software.
\end{itemize}

@@

One of the group's ideas was to create a funding program for disciplines or
other organizations to create a prize program. We would provide a framework and
a set of requirements, for example: awards to individuals; must award in 5 or 6
areas; must have a jury that includes senior/junior
domain experts and technologists (and should have objective criteria); must have the recipients awarded at a
relevant event; and must provide citations that explain why the prizes should be
awarded. Different organizations could then decide to sign up to the framework
and give out awards under this general brand. However, there was a concern that
having many organizations award prizes may reduce the impact of the prizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reproducibility \& Reuse \& Sharing} \label{sec:reproduce}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Neil led writing of this section}

%\note{\href{http://tinyurl.com/kqpe87z}{Google doc notes}}
%https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9

@@

The first group discussed ways in which reproducibility of papers could be
improved. A consensus surrounded the provision of examples: demonstrating to
others how to achieve reproducibility. Major public funding investments go into
research that heavily relies on reproducible software, hence the lack thereof
is raising major concerns.

@@

Other discussion focused on ways in which researchers themselves could improve
the reproducibility of research. One way would be to establish tracks at
conferences that subject papers to reproduction, which for those that pass would
lend them an additional badge (similar to OOPSLA). Another is simply to get more
people to use your software: for instance by outreach to high school students --
can your software be used by them? Finally, there is a role for
community-curated benchmarks to validate the performance and capabilities of
tools.

@@

An issue around the sustainability of software catalogues is that their
usefulness often depends on the domain. For instance, in the biosciences, there
is more homogeneous data and standard shared code. In areas like ecology,
code is often very specific to a problem, meaning that the level of re-use might
be at a general statistical level of abstraction, but then every research
use is highly customized.

@@

\item Paper Policies
\begin{itemize}
	\item Required sections
	\item Keywords
	\begin{itemize}
		\item Paper provides keywords to help describe software
	\end{itemize}
	\item Papers can be updated when new releases of software are made
	\begin{itemize}
		\item At no extra cost/at significantly reduced cost
	\end{itemize}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code Testing \& Code Review} \label{sec:code_testing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Dan leading writing of this section, with help from Tom Clune and Karen Cranston}

%\note{\href{http://tinyurl.com/l5t5h45}{Google doc notes}}
%https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur

@@

\item Formal tutorials at conferences (for example, a test driven development
(TDD) at a Software Developers Best Practices conference): These often require
self-learning following a tutorial, but it can be hard to find a tutorial that
matches the programming language and/or the domain of the participant.

@@

Teach testing by beginning with a (smallish) piece of code that lacks
appropriate tests and develop an exercise that involves refactoring the code
into a more presentable form through creating/deriving unit tests. Students
would extract unit-tests through reverse-engineering and/or questioning a
partner who is a developer/expert.

This led into a more general discussion about whether programming courses
should be part of the standard curriculum for science students and the
challenges of fitting additional material into university curricula. This is the
reasons that short workshops such as Software Carpentry and others exist.

@@

\subsubsection{Writing tests for scientific software}
The second major discussion thread in the group addressed the issues identified
by Software Carpentry -- the difficulties in writing tests appropriate for
scientific software: %Even among this group, while many 
What tests are
appropriate to ensure that an complex method is working correctly? When is the
result of a numerical computation `close enough' to pass a test?

Several members of the group indicated that while they valued the concept of
software testing in theory, they were unaware of how to test certain kinds of
software. Or, more importantly, they lacked relevant examples of software
testing that would be suitable for new members of their teams. As an example, the
group asked if it could produce a 100-line example of how to test neutron
transport (or other specific scientific examples) targeted at a sophomore? This
would be a demonstration of ``How do I do this right?''

@@

The fact remains that contributors also want to get credit for their
participation in the process. And the workshop organizers will want to make
sure that the workshop content and their efforts are recorded. Ideally, there
would be a service that would index the contributions to the
workshop, serving the authors, the organizers, and the larger community. 
Since there still isn't such a service today, the workshop organizers are
writing this initial report and making use of arXiv as a partial solution to
provide a record of the workshop.

@@

%\todo{feel free to add stuff here}

@@

\bibliographystyle{vancouver}
